---
title: "Monocyte Subtypes" 
author: "<h3>Author</h3>Brian M. Schilder, Bioinformatician II"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document: 
    theme: spacelab
    highlight: zenburn 
    code_folding: show 
    toc: true 
    toc_float: true
    smooth_scroll: true
    number_sections: false 
    self_contained: true
params:
  subsetGenes: "protein_coding" #FALSE
  subsetCells: 500 #FALSE
  resolution: 0.6
  resultsPath: "./Results"
  nCores: 2
  perplexity: 30
---

# Setup 

```{r setup, dpi = 600, warning=F, message=F}
# Import functions
root = "./"
source(file.path(root,"general_functions.R"))
import_parameters(params)
# load(file.path(resultsPath, "scRNAseq_results.RData"))
# resultsPath <- "Results/subsetGenes-protein_coding__subsetCells-F__Resolution-0.2__perplexity-40__nCores-4"
 
library(Seurat)
library(cowplot)
library(ggplot2)
library(dplyr) 
library(data.table) 
library(readxl) 
library(reshape2)
library(ggrepel)
library(plotly)

library(GeneOverlap) # BiocManager::install("GeneOverlap") 
library(enrichR) #BiocManager::install("enrichR")

## Upgrade to alpha (development) version of Monocle, as this version has been optimized for large datasets
if("DDRTree" %in% rownames(installed.packages())==F){
  devtools::install_github("cole-trapnell-lab/DDRTree", ref="simple-ppt-like")
}
if("graph" %in% rownames(installed.packages())==F){
 devtools::install_github("cole-trapnell-lab/L1-graph") 
}
if("reticulate" %in% rownames(installed.packages())==F){
 install.packages("reticulate")
}
if("flexclust" %in% rownames(installed.packages())==F){
 install.packages("flexclust")
}
if("mcclust" %in% rownames(installed.packages())==F){
 install.packages("mcclust")
}
library(flexclust)
library(mcclust)
library(reticulate)
if(py_module_available("umap-learn")==F){
  reticulate::py_install('umap-learn')# , pip = T, pip_ignore_installed = T # Ensure the latest version of UMAP is installed 
}
if(py_module_available("louvain")==F){
  reticulate::py_install("louvain")
  system("pip install louvain --user")
}
if("monocle" %in% rownames(installed.packages())==F){
  devtools::install_github("cole-trapnell-lab/monocle-release", ref="monocle3_alpha") 
}
if("ggrastr" %in% rownames(installed.packages())==F){
  devtools::install_github("VPetukhov/ggrastr")
}

library(monocle) # BiocManager::install("monocle") 
# BiocManager::install(c('DelayedArray', 'DelayedMatrixStats', 'org.Hs.eg.db', 'org.Mm.eg.db'))
library(org.Hs.eg.db)
library(garnett) # devtools::install_github("cole-trapnell-lab/garnett")
 
#  
if(getwd()=="/Users/schilder/Desktop/PD_scRNAseq"){
  allGenes <- F
}else{allGenes <- T}
allGenes
sessioninfo::session_info()
```

# Monocle 

## Convert and Normalize

* Convert from Seurat object to CDS object and re-normalize the data from scratch (importCDS currently only converts the raw.data, not the scale.data).
* Alternatively, you can do your normalization in Seurat and then construct a new CDS object manually with the scale.data from the Seurat object.
```{r Convert and Normalize, warning=F, message=F}
# Import Seurat obj
load(file.path(root, "Data/seurat_object_add_HTO_ids.Rdata"))
DAT <- seurat.obj  
rm(seurat.obj)
# Add metadata
metadata <- read.table(file.path(root,"Data/meta.data4.tsv"))
DAT <- AddMetaData(object = DAT, metadata = metadata)  
protDAT <- subsetBiotypes(DAT, subsetGenes = "protein_coding")
protDAT <- remove_nonmatched_metadata(protDAT, subsetCells = F)
protDAT <- FindVariableGenes(object = protDAT, mean.function = ExpMean, 
                         dispersion.function = LogVMR, 
                         selection.method ="dispersion", do.plot = T, 
                         top.genes = 2000)



# Pass TRUE if you want to see progress output on some of Monocle 3's operations
DelayedArray:::set_verbose_block_processing(TRUE)
# Passing a higher value will make some computations faster but use more memory. Adjust with caution!
options(DelayedArray.block.size=1000e6)

## Construct CDS object manually
# mDAT <- newCellDataSet(cellData = DAT@scale.data,
#                        featureData = AnnotatedDataFrame( data.frame(gene_short_name=row.names(DAT@scale.data),
#                                                                     row.names = row.names(DAT@scale.data))
#                          ),
#                        phenoData = AnnotatedDataFrame(DAT@meta.data))
# mDAT <- estimateSizeFactors(mDAT) #DESeq2?
# mDAT <- preprocessCDS(mDAT, method = "PCA", num_dim = 10, norm_method = "none")


## Construct CDS object automaticaly
### NOTE!: importCDS takes only the raw.data (not scale.data)
mDAT <- monocle::importCDS(protDAT,  import_all = T)
mDAT <- estimateSizeFactors(mDAT) #DESeq2?
mDAT <- estimateDispersions(mDAT)
# Regressing out ID will also regress out mutation status 
mDAT <- preprocessCDS(mDAT, num_dim = 20, residualModelFormulaStr = "~ nUMI + percent.mito") 
```


## Dimensionality Reduction

```{r Dimensionality Reduction}
# !Important! (allows replicability)
set.seed(1) # Monocle sets their seed to 2016 by default (Seurat sets it to 1)
# Options: c("UMAP", "tSNE", "DDRTree", "ICA", "none")

## UMAP
# monocle:::UMAP( )
mDAT <- reduceDimension(mDAT, reduction_method = 'UMAP', 
                        max_components = 3, 
                        metric="cosine",
                        n_threads =  parallel::detectCores())
## t-SNE
# mDAT <- reduceDimension(mDAT, reduction_method = 'tSNE', 
#                         perplexity = 30,
#                         scaling = F,
#                         num_threads = parallel::detectCores()) 

pretty_colors <- function(mDAT, var="Cluster", palette="custom"){
  unique_var <- unique(as.character(pData(mDAT)[var][,1]))
  if(palette=="custom"){
    col_vector_origin <- c("mediumorchid","deepskyblue","limegreen","steelblue",
                           "hotpink","turquoise", "blueviolet","mediumvioletred",
                           
                           "#db83da","#53c35d","#a546bb","#718fe8","#a469e6",
                         "#babb3d","#4f66dc","#e68821","#83b837","#d6ac3e",
                         "#7957b4","#468e36","#d347ae","#5dbf8c","#e53e76",
                         "#42c9b8","#dd454a","#3bbac6","#d5542c","#59aadc",
                         "#cf8b36","#4a61b0","#8b8927","#a24e99","#9cb36a",
                         "#ca3e87","#36815b","#b23c4e","#5c702c","#b79add",
                         "#a55620","#5076af","#e38f67","#85609c","#caa569",
                         "#9b466c","#88692c","#dd81a9","#a35545","#e08083",
                         "#17becf","#9edae5")
  } else{col_vector_origin <- RColorBrewer::brewer.pal(length(unique_var),palette)} 
  # barplot(1:length(col_vector_origin) , col=col_vector_origin) 
  col_vector <- col_vector_origin[1:length(unique_var)]
  names(col_vector) <- unique_var
  return(col_vector)
} 
```


## Clustering

* "Here __res__ represents the resolution parameter (range from 0-1) for the louvain clustering. Values between 0 and 1e-2 are good, __bigger values give you more clusters__. Default is set to be 1e-4. louvain_iter represents the number of iterations used for Louvain clustering. The clustering result gives the largest modularity score that will be used as the final clustering result. The default is 1."

### Test Clustering Hyperparameters {.tabset .tabset-fade .tabset-pills}

```{r Test Clustering Hyperparameters, results="asis"}  
plotDR <- function(resDAT, metavar="Cluster", title=""){ 
  # options(repr.plot.width = 11)
  # options(repr.plot.height = 8)
  col_vector <- pretty_colors(resDAT, var=metavar) 
  p <- plot_cell_clusters(resDAT,
                     color_by = metavar,
                     cell_size = 0.8,
                     show_group_id = T)  + 
    scale_color_manual(values = col_vector) +
    theme(legend.text=element_text(size=6)) +
    theme(legend.position="right")+ 
    labs(title = paste(title)) 
  print(p)
}

test_hyperparameters <- function(mDAT, resolutions=seq(0, 1e-4,length.out=6),
                                 Ks=seq(10,60,length.out=6), 
                                 iter_var="k"){
  if(iter_var=="resolution"){
    for(res in resolutions){
      res_title <- paste("Resolution =",res)
      cat("\n")
      cat("####",res_title,"\n")
      try({ 
       resDAT <- clusterCells(mDAT, res=res,# k=43,
                          method = "louvain",# densityPeak
                          louvain_iter = 1,
                          verbose = F, 
                          clustering_genes = clustering_genes,
                          cores = parallel::detectCores())
      plotDR(resDAT, title=res_title)  
      }) 
      cat("\n")
    }  
  }else{
    for(k in Ks){
      k_title <- paste("K =",k)
      cat("\n")
      cat("###",k_title,"\n")
      try({ 
       kDAT <- clusterCells(mDAT, res=8.888889e-05, k=k,
                          method = "louvain",# densityPeak
                          louvain_iter = 1,
                          verbose = F, 
                          clustering_genes = clustering_genes,
                          cores = parallel::detectCores())
      plotDR(kDAT, title=k_title)  
      })
      cat("\n")
    }   
  } 
}
test_hyperparameters(mDAT, iter_var="resolution")
# test_hyperparameters(mDAT, iter_var="k")
```

### Final Clustering Selection

```{r Final Clustering Selection} 
# Select final resolution
## Use ONLY the 1000 most variable genes to do clustering (otherwise there's too much noise)
clustering_genes <- intersect(mDAT@featureData@data$gene_short_name, protDAT@var.genes)[1:1000]
louvain_res <- 1e-04 # 1e-04 is the default
mDAT <- clusterCells(mDAT, res=louvain_res,
                        method = "louvain",
                        louvain_iter = 1, 
                        clustering_genes = clustering_genes,
                        verbose = T, cores = parallel::detectCores())
# mDAT <- clusterCells(mDAT, num_clusters = 4,
#                         method = "densityPeak",    
#                         verbose = F,  
#                         cores = parallel::detectCores())
plotDR(mDAT, "Cluster")
plot_cell_clusters(mDAT, color_by = "dx", cell_size = .8)
plot_cell_clusters(mDAT, color_by = "mut", cell_size = .8) 
# CLUSTERING FROM SEURAT
# plotDR(mDAT, "post_clustering")
```


## Identify Cell Types with Garnett

```{r Identify Cell Types with Garnett} 
# generate size factors for normalization later
# Get pre-trained PBMC classifer 
load(file.path(root, "Data/Garnett/hsPBMC")) # Download from: https://cole-trapnell-lab.github.io/garnett/classifiers/hsPBMC  

mDAT <- garnett::classify_cells(mDAT, 
                           classifier = hsPBMC,
                           db = org.Hs.eg.db,
                           cluster_extend = TRUE,
                           cds_gene_id_type = "SYMBOL")
table(pData(mDAT)$cell_type)
cell_summary <- table(pData(mDAT)$cluster_ext_type) 
cell_summary
# Get feature genes for each cell type
feature_genes <- garnett::get_feature_genes(classifier = hsPBMC,
                                   node = "root",
                                   db = org.Hs.eg.db,
                                   convert_ids = F)
head(feature_genes)  
# If a cell type is called less than n times, re-categorize it as unknown
mDAT$cluster_ext_type_filt <- ifelse(mDAT$cluster_ext_type %in% names(cell_summary[cell_summary<50]), 
                                     "Unknown", mDAT$cluster_ext_type)

plot_cell_clusters(mDAT, color_by ="cell_type", cell_size = .8) +  facet_wrap(~dx)
plot_cell_clusters(mDAT, color_by ="cluster_ext_type", cell_size = .8)  +  facet_wrap(~dx)
plot_cell_clusters(mDAT, color_by ="cluster_ext_type_filt", cell_size = .8) +  facet_wrap(~dx)
```

## Gene Expression

### Monocyte Markers

```{r Monocyte Markers}
plot_cell_clusters(mDAT, markers = c("CD14","FCGR3A"), cell_size = 0.8)
```

### PD Genes

```{r PD Genes}
plot_cell_clusters(mDAT, markers = c("LRRK2","GBA"), cell_size = 0.8)
```
 
## Selected Genes

```{r Selected Genes}
plot_cell_clusters(mDAT, markers = c("MS4A4A","MS4A6A"), cell_size = 0.8) 
```

## Heatmap
```{r Heatmap}
plot_markers_cluster(mDAT, markers = protDAT@var.genes[1:50], minimal_cluster_fraction = 0.05)
```


## Pseudotime

```{r Pseudotime}  
# subsetCDS(mDAT, )
mDAT <- partitionCells(mDAT)
mDAT <- learnGraph(mDAT, RGE_method = 'SimplePPT',
                   n_threads =  parallel::detectCores())

plot_cell_trajectory(mDAT, color_by = "Cluster", cell_size = .01) 
plot_cell_trajectory(mDAT, color_by = "dx", cell_size = .01) 
plot_cell_trajectory(mDAT, color_by = "mut", cell_size = .01) 
plot_cell_trajectory(mDAT, color_by = "cluster_ext_type_filt", cell_size = .01) 
 
# dir.create(file.path(resultsPath, "pseudotime"), showWarnings = F)
# plot_3d_cell_trajectory(mDAT,
#                         color_by="cell_type",
#                         webGL_filename= file.path(resultsPath, "pseudotime/pseudotime_cellType.html"),
#                         show_backbone=TRUE,
#                         useNULL_GLdev=TRUE)
# plot_3d_cell_trajectory(mDAT, markers = c('FCGR3A'),
#                         webGL_filename=file.path(resultsPath, "pseudotime/pseudotime_FCGR3A.html"),
#                         show_backbone=TRUE,
#                         useNULL_GLdev=TRUE)
```

## Differential Expression: Monocle

* [Tutorial](http://cole-trapnell-lab.github.io/monocle-release/docs/#differential-expression-analysis)
* qval = FDR
*"The first of the two models is called the full model. This model is essentially a way of predicting the expression value of each gene in a given cell knowing only whether that cell is a fibroblast or a myoblast. The second model, called the reduced model, does the same thing, but it doesn't know the CellType for each cell. It has to come up with a reasonable prediction of the expression value for the gene that will be used for all the cells...The question Monocle must answer for each gene is how much better the full model's prediction is than the reduced model's."
```{r Differential Expression: Monocle, eval=F} 
mDAT_sub <- mDAT[protDAT@var.genes[1:50], mDAT@phenoData$Cluster %in% c(1,2)]

# At ~1min/100 genes, this DGE method will take ~2.5hours to run for 14827 genes 
DEG_df <- differentialGeneTest(mDAT_sub, verbose = T,
                               fullModelFormulaStr = "~Cluster",
                               cores = parallel::detectCores())

# Plot
mDAT_sub_genes <- mDAT_sub[row.names(DEG_df)[1:4],]
plot_genes_jitter(mDAT_sub_genes, grouping = "Cluster", ncol= 4, color_by = "dx", plot_trend = T)
```


# Save Results

```{r Save Results}  
save.image(file.path(resultsPath, "scRNAseq_results.RData"))
```


